---
title: "Extract transcript sequences"
author: "Sofia Salazar"
date: "2025-03-27"
output:
  pdf_document: default
  html_document: default
---

## Required packages

```{r eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

BiocManager::install("GenomicFeatures")
BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
BiocManager::install("TxDb.Hsapiens.UCSC.hg38.knownGene")
```


```{r warning=FALSE}
suppressPackageStartupMessages(library(GenomicFeatures))
suppressPackageStartupMessages(library(BSgenome.Hsapiens.UCSC.hg38))
suppressPackageStartupMessages(library(TxDb.Hsapiens.UCSC.hg38.knownGene))
```

## TxDB object

Create a TxDb object 

```{r}
hg38_txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
hg38_txdb
```

## Extract GRanges object

Extracting a GRanges object with genomic features from the TxDb object

**Transcripts**

I think the following function extracts transcripts from the genome in the format of a GRanges object.

```{r eval=FALSE}
?transcripts
```

```{r}
GR <- transcripts(hg38_txdb)
GR[10:13]
```
```{r}
transcripts_df <- as.data.frame(GR)
length(transcripts_df$tx_id)
```

Extracting transcripts from a given transcript ID

```{r}
transcripts(hg38_txdb, filter = list(
  tx_name = "ENST00000456328.2"
))
```

**Exons**

Extract exons for each transcripts
```{r}
exons_tx <- exonsBy(hg38_txdb, by = "tx")
length(exons_tx)
```

```{r}
names(exons_tx)[10:13]
```

Here the names of the exons_tx column correspond to the tx_id from the transcripts

```{r}
tx_ids <- names(exons_tx)
head(select(hg38_txdb, keys=tx_ids, columns="TXNAME", keytype="TXID"))
```

## Extract sequences

Here, genes are queried by ENTREZ IDs, so I load a library to query the ENTREZ IDs for a couple of hemoglobin genes


```{r}
library(org.Hs.eg.db)
my_genes <- c("HBA1", "HBA2", "HBB")
my_entrez <- select(org.Hs.eg.db, 
       keys = my_genes,
       columns = "ENTREZID", "SYMBOL",
       keytype = "SYMBOL")
my_entrez$ENTREZID
```

**Query the transcripts for each gene**

Here, each `name(txbygene)` has the transcripts for each gene

```{r}
txbygene <- transcriptsBy(hg38_txdb, "gene")[my_entrez$ENTREZID]
txbygene[1:3]
```

**Extract UTR and coding regions**

With `use.names` each `names(cds)` or `names(threeUTR)` will be a transcript ID

```{r}
cds <- cdsBy(hg38_txdb, by="tx", use.names=TRUE)
threeUTR <- threeUTRsByTranscript(hg38_txdb, use.names=TRUE)
fiveUTR <- fiveUTRsByTranscript(hg38_txdb, use.names=TRUE)
```

```{r}
head(names(fiveUTR))
```

Extract transcript IDs from each gene

```{r}
map <- relist(unlist(txbygene, use.names=T)$tx_name, txbygene)
map
```

```{r}
my_transcripts <- unlist(map, use.names=F)
head(my_transcripts)
```


```{r}
my_coding_regions <- cds[names(cds) %in% my_transcripts]
my_coding_regions
```

**Extract sequences**

```{r}
genome <- BSgenome.Hsapiens.UCSC.hg38
my_coding_seqs <- extractTranscriptSeqs(genome, my_coding_regions)
my_coding_seqs
```

**Translate**

```{r}
translate(my_coding_seqs)
```

## Try on HBB

```{r}
# Step 1: Get Entrez ID for HBB
hbb_entrez <- AnnotationDbi::select(org.Hs.eg.db, 
                                    keys = "HBB", 
                                    keytype = "SYMBOL", 
                                    columns = "ENTREZID")$ENTREZID
hbb_transcripts <- transcripts(hg38_txdb, filter = list(gene_id = hbb_entrez))

# View the result
hbb_transcripts
```

